%\section{Introduction}

%    a. chirality - same story as my Internal Seminar.
%    b. Drug thalidomide example
%    c. Chiroptical spectroscopy
%    d. Theory as a tool!
%    e. Some history of OR calculations (Harley's thesis)
%    f. TD-DFT not enough in most cases, CC needed (known problems of DFT)
%    g. Challenges of CC - (mention about MD as well.)
%    h. My work comes in here!
%	i. local correlatin - which chapter discusses what... end by conclusion..
\section{Chirality}
Lord Kelvin was the first person to use the word `chiral' in his lecture 
at Oxford University in 1894: ``I call any geometrical figure, or group of points, 
`chiral', and say that it has chirality if its image in a plane mirror, ideally realized, 
cannot be brought to coincide with itself"\cite{Kelvin}. 
\\
There are different ways in which a molecule 
could be chiral. The most common example being the presence of a chiral center 
like a carbon or nitrogen atom with four different substituents. Axial chirality, usually 
possessed by molecules with cumulated double bonds like allenes and chirality planes 
found in compounds like paracyclophanes are some other popular examples. \\
A chiral molecule and its mirror image are known as enantiomers. Although the
enantiomers have idential physical properties like boiling and melting points, their
interactions with a chiral environment can be quite different. For example,
while the (R) enantiomer of the compound limonene smells like lemons and oranges, the (S) 
enantiomer on the other hand smells like turpentine. Thus, the interaction of the (S) enantiomer 
with the chiral molecules constituting the olfactory receptors responsible for the sensation of smell 
is totally different from its fellow enantiomer. The enantiomers also interact differently with 
the left- and right-hand circularly polarized (LCP and RCP) light in processes of absorption, 
refraction, etc. A chiral molecule posesses the properties of circular dichroism (CD) and 
circular birefringence, i.e. it has different absorption coefficents and refractive indices 
for the LCP and RCP light respectively. As a result of circular birefringence, the plane of 
polarization of a plane-polarized light in a chiral medium gets rotated/shifted by an amount known as 
the optical rotation of the medium. An enantiomeric pair of chiral molecules has equal magnitudes of 
optical rotation but in opposite directions.\\
Since almost every part of the human body is composed of chiral molecules, its unsurprising that
more than 60\% of the pharmaceutical drugs are chiral as well. One of the most 
important aspect of the drug-design research is to be able to define the absolute configuration 
(AC) of these drug molecules in order to study their behaviour in biological systems as 
many biological activities are only associated with one specific AC. A very tragic example
in this regard is that of the drug thalidomide, which was originally prescribed to pregnant women in 
Europe in 1950s to cure morning sickness but resulted in birth-defects in thousands of 
infants. It was found through a study done on rodents that the (R) enantiomer of thalidomide is indeed 
a sedative but the (S) enantiomer is a teratogen\cite{Crawford06}. Furthemore, these enantiomers were 
found to rapidly interconvert in vivo in humans because of which separating 
these two forms before use doesn't help at all.
\\
The first step in experimental determination of AC either involves the separation of 
a racemic mixture (equal concentrations of both the enantiomers) into individual enantiomers
using chiral resolution methods like chiral column chromatography or a selective synthesis
of a given enantiomer using the process of asymmetric synthesis. X-ray crystallography would then be the 
most popular procedure to determine AC directly, provided one can crystallize the molecule first, which usually 
requires the presence of heavy atom(s). Alternatively, one can measure the distinct responses of 
the enantiomers in absorption using chiroptical spectroscopic techniques like electronic and vibrational 
CD (ECD and VCD) or can measure the optical rotation of the chiral compounds in both liquid 
and gas phase using polarimetry and cavity ring down spectrometry techniques\cite{} respectively.
These measurement are usually followed by a comparison against some suitable reference chiral
substrate. In this regard, accurate {\em ab initio} methods can help validate these experimental measurements.
%However, in cases of ambiguities regarding appropriate references, accurate {\em ab initio} methods can come to the rescue and 
%help validate these experimental measurements. 
Furthermore, designing an asymmteric synthesis might not be straightforward for many chiral systems, 
especially for the ones with large number of stereocenters. In such cases, theory can even take the 
lead and predict the corresponding AC. In other words, theoretical calculations of these chiroptical 
properties could serve as a very useful computational tool for modern organic chemists. In particular, 
accurate optical rotation calculations are highly desired as the experimental techniques for measuring 
this property are fairly stable and robust in both gas and liquid phases.
\\
%Theoretical and computational chemistry has grown by leaps and bounds over the
%past several decades. {\em ab initio} quantum chemical methods can now accurately
%predict a variety of molecular properties. More recently, these methods have
%been extended to calculate chiroptical
%properties like optical rotation. Theoretical calculations of optical rotation
%can be very useful in solving experimental chemical problems.  Specifically, they
%can provide a computational tool to modern organic chemists to determine
%the absolute stereochemical configuration of a chiral compound, thus alleviating the experimental difficulties
%involved.\\ 
\section{{\em Ab Initio} Optical Rotation Calculations}
Rosenfeld developed a quantum mechanical recipe for calculating optical rotations in the year
1929\cite{Rosenfeld29}. He showed that in the presence of an external electromagnetic field,
the induced dipole moment in a molecule can be written as:
\begin{equation}\langle\vec{\mu}\rangle = \bm{\alpha}\vec{E} + \frac{1}{\omega}\textbf{G}^\prime\frac{\partial\vec{B}}{\partial t}
\end{equation} where $\vec{E} $ and $\vec{B}$ are the time dependent electric and
magnetic field vectors respectively, $\omega$ is the frequency of the field, $\alpha$ is the electric dipole
polarizability tensor and the $\textbf{G}^\prime$ tensor, also known as the Rosenfeld tensor is associated 
with the property of optical rotation. Invoking the time-dependent Schr\"odinger's equation in conjunction 
with perturbation theory, expanding the expectation value of the electric dipole operator in orders of perturbation 
and identifying the term linear in $\vec{E}$ and $\frac{\partial\vec{B}}{\partial t}$, 
for exact states, the $\bm{\alpha}$ and $\textbf{G}^\prime$ tensors look like,
\\
\begin{equation}
\begin{split}
& \bm{\alpha}(\omega) = -\frac{2}{\hbar} \sum_{n \neq 0}\frac{\omega_{no}\bra
{\psi_o}\vec{\mu}\ket{\psi_n}\bra{\psi_n}\vec{\mu}\ket{\psi_o}}{{\omega_{no}}^2 - {\omega}^2 }\\
& \textbf{G}^{\prime}(\omega) = -\frac{2}{\hbar} \text{Im}\sum_{n \neq 0}\frac{\omega \bra
{\psi_o}\vec{\mu}\ket{\psi_n}\bra{\psi_n}\vec{m}\ket{\psi_o}}{{\omega_{no}}^2 - {\omega}^2 }.\\
\end{split}
\end{equation}
\\
Here, $\vec{\mu}$ and $\vec{m}$ are electric and magnetic dipole operators, 
$\vec{\mu} = \sum\limits_i q_i \vec{{r}_i}$ and $\vec{m} = \sum\limits_i \frac{q_i}{2m_i} \vec{{r}_i} \times \vec{{p}_i}$, $\omega$ is the frequency of light, $\omega_{no}$ is the excitation energy of the state $\psi_n$.
%\cite{CrawfordTamJPA07}. 
Im means the imaginary part of the equation and the summation runs over all the excited
states $\psi_n$. Specific rotation, ${\lbrack\alpha\rbrack}_\omega$ is related to the trace of the Rosenfeld tensor 
normalized by pathlength and concentration and is usually reported in the units of deg dm$^{\text{-1}}$(g/mL)$^{\text{-1}}$,\cite{Crawford06}
\\
\begin{equation}
{\lbrack\alpha\rbrack}_{\omega} = \frac{(72.0 \times 10^6){\hbar}^2 N_A\;\omega}{c^2{m_e}^2 M} \times \left[ \frac{1}{3}Tr(\textbf{G}^\prime)\right]
\end{equation}
\\
Here, $\textbf{G}^\prime$ and $\omega$ are in atomic units, c is the speed of light (m/s), m$_{\text{e}}$ is the 
mass of electron (kg), M is the molecular mass (amu) and $N_A$ is Avogadro's number. Calculating the 
$\textbf{G}^\prime$ tensor using equation (1.2) requires explicit calculations of a large number of 
excited states which is of course computationally prohibitive. A more practical approach is to invoke the response formalism
\cite{Kobayashi94,Koch90}, which casts these tensors as response functions,
\begin{equation}
\begin{split}
&\bm{\alpha}(\omega) = -\langle\langle\vec{\mu};\vec{\mu}\rangle\rangle\\
&\textbf{G}^{\prime}(\omega) = \text{Im}\langle\langle\vec{\mu};\vec{m}\rangle\rangle.\\
\end{split}
\end{equation} 
The response theory focuses on the perturbation of the ground state wavefunction in the presence of an external 
field, avoiding explicit calculations of excited states. Polavarapu was the first to use this theory to calculate 
{\em ab initio} specific rotations using the time dependent Hartree Fock (TDHF) method \cite{Polavarapu96}. 
The signs of the calculated values matched with that of the experiment for most of his structures, while their magnitudes 
differed generally by a factor of two. Inspired by Polavarapu's work, Cheeseman et al. \cite{Cheeseman00,Stephens01} 
included correlation effects in these calculations by using TD-DFT theory. Using large basis sets with diffuse functions like 
aug-cc-pVDZ and aug-cc-pVTZ,\cite{Dunning89,Kendall92,Woon94} they were able to match the experimental values very closely, with a 
deviation of 20-25 degrees for a set of 28 chiral molecules. Ruud {\em et al.} extended the calculations to the coupled cluster (CC)
level using coupled cluster response theory developed by Koch and J{\o}rgensen\cite{Koch90}, and obtained very 
promising results\cite{Ruud03}. Unsurprisingly, these simulations have been used to determine the AC of a large number of 
chiral molecules over the years\cite{Kondru99}. However, for a truly robust and reliable description of these response 
properties, one needs to take into account, a number of challenges associated with such simulations.
%However, all these calculations were done on single isolated molecules and thus, ideally, they should only be 
%expected to match vapor phase optical rotation (OR) values. 
\subsection{Challenges}
One of the major challenges faced by the coupled cluster respone theory is that of gauge and origin invariance.
The origin dependence while using length-gauge representation is a consequence of the non-variational nature of the 
coupled cluster wavefunction. Unlike HF and DFT methods, properties such as optical rotation computed via CCLR 
are not origin independent even at the complete basis set limit, and the dependence cannot be resolved using GIAOs.6 
The modified velocity gauge formalism for the electric dipole operator has been developed to help address this 
issue. The velocity-gauge formalism, ⟨⟨⃗p, L⃗ ⟩⟩ω , allows for the trace of the property tensor to be origin 
invariant. Here ⃗p and L⃗ are the electronic momentum and angular momentum operators, respectively, and 
⟨⟨, ⟩⟩ω represents the response function evaluated at frequency ω. However, without a complete basis
set, using the velocity gauge formalism results in unphysical behavior at zero frequency. To achieve the 
appropriate behavior, the value for the velocity gauge representation at the static limit (at zero frequency) 
is subtracted from the original expression, as shown in Equation 1.1. 13
⟨⟨⃗p, L⃗ ⟩⟩ω → ⟨⟨⃗p, L⃗ ⟩⟩ω − ⟨⟨⃗p, L⃗ ⟩⟩0 (1.1) Alternatively, the underlying response theory can be reformulated to incorporate 
orbital rotation parameters explicitly to achieve origin invariance.14,15\\
%Careful consideration of numerous factors, including electron correlation, basis-set completeness, gauge representation, 
%zero-point vibrational motion, temper- ature, conformational flexibility, and so forth, have aided the development of advanced quantum chemical calculations.
%which can be measured by using cavity ring down spectrometry, thanks to the pioneering 
%works of Vacaaro and co-workers.\cite{} 
%\subsubsection{Electron Correlation}
%Mary Tam's DFT work maybe - need for CC methods. Crawford08?
%\subsubsection{Modeling Solvation}
Since, most of the optical rotation measurements are carried out in liquid phase or solutions,
response theory needs to be combined with accurate solvation models in order to match these
experimental measurements. While a quantum mechanical description of both the solute and solvent molecules would be ideal, 
however, for the sake of computational practicality, most solvation models try to reduce the degrees of freedom of
the solvent molecules without compromising the quality of description of the solvent-solute and the solvent-solvent
interactions.\\
In this regard, molecular mechanics (MM) based models, broadly known as QM/MM methods, usually represent the
solvent in the form of a classical potential by replacing them with point charges, polarizabilities, etc.,\cite{}
which polarizes the charge distribution of the solute resuting in a modification of the solute's potential which
again back-polarizes the charge distribution of the solvent and so on. Thus, the final one-electron solvent potential is obtained in a
self-consistent procedure and is added to the solute's Hamiltonian which is then followed by response calculations. The fluctuating
charge model (FQ)\cite{}, the drude oscillator model\cite{} and effective fragmentation potential (EFP)\cite{} are a few 
examples.\\
Quantum continuum solvation models on the other hand place the solute in a cavity surrounded by a continuous
dielectric medium representing the solvent. A majority of these continuum models use the
concept of apparent surface charges\cite{}, where polarization charges are introduced on the surface of the 
cavity, to capture the response of the medium to the charge distribution of the solute which is again, 
solved in a self-consistent manner as described above. The polarizable continuum model (PCM) \cite{} and conductor-like 
screening model (COSMO) \cite{} are some popular examples.\\
One of the earliest works employing a solvation model for studying specific rotations is that of 
Mennucci and co-workers in 2002\cite{Mennucci02}, where they used DFT and PCM for calculating 
specific rotations of a set of seven chiral compounds in the presence of seven different 
solvents. While they obtained promising results for cyclohexane, acetone, methanol and 
acetonitrile, the model performed rather poorly for carbon tetrachloride, benzene, and chloroform
due to the failure of the PCM in capturing the dominant nonelectrostatic solute-solvent interactions 
in the latter. A few years later, Pecul {\em et al.} \cite{} used the same model (DFT-PCM) to study the effect
of solvent on the electronic circular dichroism spectra and obtained promising results for valence-only transitions. 
However, this approach failed to describe the Rydberg transitions and specific solute-solvent interactions like hydrogen bonding accurately.\\
A direct way of accounting for the dynamic solvent effects is to use an explicit {\em ab initio} solvation model where
both solute and solvent are treated quantum mechanically and final specific rotations are obtained by averaging their
values over a series of configurations usually generated through molecular dynamics (MD) simulations. Beratan and co-workers 
used this model with DFT to calculate specific rotations of methyloxirane in water\cite{} and benzene\cite{} and were 
able to match the experimental values qualitatively. Barone {\em et al.}\cite{} were also able to reproduce the experimental optical
rotatory curve of methyloxirane (qualitatively) in water by using a DFT/FQ/PCM approach. However, it remains to be seen if these 
models perform equally well in other solvents and with more accurate theories like coupled cluster as DFT has been shown to agree 
with gas-phase experimental data of methyloxirane only fortuitously\cite{}.\\
Kongsted and co-workers were the first to use a solvation model within the context of coupled cluster response theory,
where they placed the solute in a sperical cavity surrounded by a dielectric medium. However, this approach 
could not match experimental measurements even qualitatively and gave a wrong sign of the specific rotation for ({\em S})-methyloxirane 
in cyclohexane at 355 nm, even after incorporating accurate electron correlation theories\cite{}. Similar behavior was observed 
with PCM as well, as demonstrated by Caricato\cite{}.\\
Another solvation model that deserves a mention here is the frozen density embedding scheme, introduced by Wesolowski and 
Warshel\cite{}. In this model, {\em ab initio} electronic densities (or potentials) describing the environment (or solvent) 
can be obtained (or decoupled) from a DFT calculation on the full system. The potential can then be added to the Hamiltonian
of the solute, followed by either a DFT (DFT-in-DFT) or a wavefunction (WFT-in-DFT) calculation\cite{}. Neugebauer
was the first to use the DFT-in-DFT FDE model to study the response properties of solvated molecular clusters\cite{}. 
Crawford and co-workers\cite{} applied the more accurate (but more expensive) WFT-in-DFT FDE model with coupled cluster theory 
to study chiroptical properties of several ``micro-solvated" structures of ({\em P})-dimethylallene in water. However, the
FDE potentials resulted in very small perturbations in the solute's electronic distribution compared to an explicit 
calculation. Furthermore, it failed to reproduce the correct sign of the specific rotations of some of 
the structures. The reason behind the errors was attributed to the failure of the FDE model in capturing
the response of the solvent to external field, which can have significant contributions for these properties.\\
%Since rotation is a dynamic property, one needs to average OR values over several 
%configurations or snapshots (obtained by molecular dynamics (MD) simulations) to be able to match experimental 
%values. 
%This restricts the application of CCLR to very small systems as each CC calculation exhibits
%a higher-order polynomial scaling.
%\subsubsection{Gauge and Origin Invariance}
%Mention coleman's and taylor's basis set work 
%\subsubsection{Scaling}
Other challenges: Basis set, scaling of CC linear response calculations, Conformational averaging, Vibrational corrections,
temperature effects, etc.
\subsubsection{Current Work}
The only robust way of accounting for solvent effects seems to be including the solvent molecules explicitly 
in the QM calculations. Furthermore, CC theory because it is more reliable, accurate and systematicaly 
improvable compared to DFT.
An obvious way to make these compuations practical is to seek the help of CC reduced-scaling 
techniques like the pair natural orbitals (PNO) approach\cite{NeeseCCSD09,Neese09} which has been really successful in describing energetics of 
large molecular systems. However, work towards the extension of these methods to calculate chiroptical properties has been 
fairly limited\cite{Friedrich15,Gauss00,Korona04,McAlexander12,Russ04,Russ08}. McAlexander and Crawford in their recent 
work\cite{McAlexander15:LRCC} demonstrated that the regular PNO approach while calculating CC response properties 
suffers from slow convergence towards the full canonical result. \\
The main focus of this work is identifying the 
reasons behind the sub-optimal peformance of such approaches for optical rotations. Subsequently, we propose novel modifications
which can improve the performance of these schemes significantly. The necessary equations required to calculate dynamic polarizabilities 
and specific rotations within the CC linear response formalism are derived in Chapter 2. 
Chapter 3 investigates the performance of the natural orbitals (FVNO scheme) for CC dynamic polarizabilites.
A new scheme called FVNO++ is developed in Chapter 4 which reduces the errors associated with the regular FVNO scheme
significantly for both polarizabilities and rotations. Chapter 5 extends the FVNO++ (a reduced-prefactor) approach to the reduced-scaling 
domain by implementing what we call as the PNO++ method. Finally, concluding remarks are given in Chapter 6.
