\begin{enumerate}
\item {\bf A. Kumar} and T. D. Crawford. Perturbed Natural Orbitals for Coupled-Cluster Linear-Response Theory. 
{\bf 2018}, {\em manuscript in preparation}.
\item {\bf A. Kumar} and T. D. Crawford. Perturbed Pair Natural Orbitals for Coupled-Cluster Linear-Response Theory. 
{\bf 2018}, {\em manuscript in preparation}.
\item D. G. A. Smith, L. A. Burns, D. A. Sirianni, D. R. Nascimento, {\bf A. Kumar}, A. M. James, J. B. Schriber, T. Zhang, B. Zhang, 
A. S. Abbott, E. J. Berquist, M. H. Lechner, L. A. Cunha, A. G. Heide, J. M. Waldrop, T. Y. Takeshita, A. Alenaizan, D. Neuhauser,
R. A. King, A. C. Simmonett, J. M. Turney, H. F. Schaefer, F. A. Evangelista, A. E. DePrince, 
T. D. Crawford, K. Patkowski and C. D. Sherrill. Psi4NumPy: An Interactive Quantum Chemistry Programming Environment for 
Reference Implementations and Rapid Development. {\em J. Chem. Theory Comput.}, {\bf 2018}, doi: 10.1021/acs.jctc.8b00286.
\item R. M. Parrish, L. A. Burns, D. G. A. Smith, A. C. Simmonett, A. E. DePrince, E. G. Hohenstein, U. Bozkaya, A. Y. Sokolov, 
R. D. Remigio, R. M. Richard, J. F. Gonthier, A. M. James, H. R. McAlexander, {\bf A. Kumar}, M. Saitow, X. Wang, B. P. Pritchard, 
P. Verma, H. F. Schaefer, K. Patkowski, R. A. King, E. F. Valeev, F. A. Evangelista, J. M. Turney, 
T. D. Crawford, and C. D. Sherrill. Psi4 1.1: An open-source electronic structure program emphasizing automation, advanced libraries, 
and interoperability. {\em J. Chem. Theory Comput.}, {\bf 2017}, {\em 13}(7), pp 3185-3197.
\item {\bf A. Kumar} and T. D. Crawford, {\em J. Phys. Chem. A}, Frozen Virtual Natural Orbitals for Coupled-Cluster 
Linear-Response Theory. {\bf 2017}, {\em 121}(3), pp 708-716.
\item T. D. Crawford, {\bf A. Kumar}, K. P. Hannon, S. H{\"o}fener and L. Visscher. Frozen-Density Embedding 
Potentials and Chiroptical Properties. {\em J. Chem. Theory Comput.}, {\bf 2015}, {\em 11}(11), pp 5305-5315.
\item J. Friedrich, H. R. McAlexander, {\bf A. Kumar} and T. D. Crawford. Incremental evaluation of coupled cluster 
dipole polarizabilities. {\em Phys. Chem. Chem. Phys.}, {\bf 2015}, 17, pp 14284-14296.
\end{enumerate}
